---
title: "Tutrorial R Basics"
author: ''
date: "nov9-2022"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    highlight: tango
---

```{r setup, include=FALSE}

# To set global options that apply to every chunk in your file
knitr::opts_chunk$set(echo = T, error = F, message = F, warning = F)

```

```{r libraries, include=F, message=T}

# package for convenience functions (e.g. ggplot2, dplyr, etc.)
library(tidyverse)

```

# <span style = "color:firebrick">extra</span>

**Set the theme for RStudio**  Tools → Global options → Appearance

**html reference** [referenfe for html](https://bookdown.org/yihui/rmarkdown/html-document.html)

**chunk options** [R chunk options](https://rmarkdown.rstudio.com/lesson-3.html)

```{r, warning=T}
# This Chunk is to understand what is 'warning' and how to mask it in the html knitted file
# change the value of 'warning' option in this chunk to see the effect

# define character vector
x <- c('1', '2', '3', NA, '4', 'Hey')

# convert to numeric vector
x_num <- as.numeric(x)
x_num
```

**Rstudio keyboard shortcuts:** Tool → Keyboard Shortcuts help → see all shortcuts 

- Insert chunk:	Ctrl+Alt+I(win) Command+Option+I(mac)
- Insert assignment operator: 	Alt+-(win) Option+-(mac)
- Comment/uncomment current selection: 	Ctrl+Shift+C(win) 	Command+Shift+C(mac)
- Insert pipe operator: Ctrl+Shift+M(win) 	Command+Shift+M(mac)


# <span style = "color:firebrick">Exercise 1:</span> matrix

Create this matrix:

```{r, eval = F}

1  3
5  3
9  3

```

**Solution 1**

create an empty matrix and fill it one by one! 
```{r}
e_matrix <- matrix( , nrow=3, ncol=2)
e_matrix


e_matrix[1,] <- c(1,3) # indexing, first row 
e_matrix

```

**Solution 2**

create a vector and turn it into matrix 
```{r}
v_1 <- c(seq(from = 1, to = 9, by = 4))
v_2 <- c(rep(3,3))

m_1 <- matrix(data = c(v_1,v_2), ncol = 2, byrow = FALSE)
m_1

```

```{r}
v_3 <- c(1,3,5,3,9,3)
matrix(v_3, ncol = 2, byrow = TRUE)
```
```{r}
v_4 <- c(1,5,9,3,3,3)
matrix(v_4, ncol = 2, byrow = FALSE)
```

**solution 3**

creat the vector and bind them together using 'cbind'
```{r}
# one to bind them 
m_2 <-  cbind(v_1,v_2)
m_2

```

**another solution + Error**

```{r}
e_matrix <- matrix( , nrow=3, ncol=2)
mat1.data <- c(seq(from = 1, to = 12, by = 4))
mat2.data <- c(rep((3),3))

# ERROR / Change nrow to 5
# make sense of the error and why the result is this way when nrow=6
e_matrix <- matrix(nrow=6, ncol=2) 
e_matrix[,1] <- mat1.data
e_matrix[,2] <- mat2.data
e_matrix
```

**solution 4**

```{r}
# dim function
mat_1 <- c(1,5,9,3,3,3) 
dim(mat_1) <- c(3,2)
mat_1
```

# <span style = "color:firebrick">Exercise 2:</span> df & tibble
There are few difference between dataframe and tibble. This difference is also valid for read.csv() and read_csv(). First one load the csv file as 'dataframe', the second one as 'tibble'.

few difference are as follows:

* printing (see below)
* access to partial provided columns (see below)
* Tibble load faster

other differences could be found in the links below:

[Tibble vs. DataFrame](https://jtr13.github.io/cc21fall1/tibble-vs.-dataframe.html)

There are many data sets available in base r 

```{r}
# list of datasets 

# data()
# library(help = "datasets")
ls("package:datasets") # list the datasets
```

```{r}
# dataframe
iris_df <- iris

# tibble
iris_tib <- as_tibble(iris)

```

```{r}
# 1.prining (dataframe prints the whole dataset but tibble only prints the top10)
iris_df
iris_tib
```

```{r}
# 2 partial col name is acceptable as long as it is unique in dataframe but tibble....
iris_df$Petal.Len
iris_tib$Petal.Leng
```

```{r}
# different ways of slicing the tibble/dataframe are the same,
# you may see the different syntax below

# iris_tib$Petal.Length
# iris_tib[3]
iris_tib["Petal.Length"] 

```


# <span style = "color:firebrick">Exercise 3:</span> functions & piping 
```{r}
# defining function
add_five <- function(x){
  x + 5
}

# anonymous function
sapply(1:20, function(i) i + 5)

# using the defined function
sapply(1:20, add_five)

# usage of anonymous function within another function
purrr::map_dbl(1:20, ~.x + 5)
purrr::map_int(1:20, ~.x + 5L) # L indicates the number's type to be integer 

```

**another example of using anonymous function**
```{r}
x <- sample(1:6, 12, replace=TRUE)
x
mat <- matrix(x, nrow=3)
mat
apply(mat, 1, function(x) { seq(min(x), max(x)) })
```

Flavors of piping:

There are two flavors of piping that you will be seeing in this course:

- The magrittr package infix operator %>% and 
- Base R native pipe operator |> in version 4.1.0

To change to Base R pipe keyboard shortcut:
Tools → Global Options → code → use native pipe

```{r}
# Piping and function
1:20 %>% 
  (function(i) {i + 5})

1:20 |>
  (function(i){i + 5})()
```


# difference between ( and [ 

```{r}
# vector
x1<-5:12
# length of the vector
length(x1[x1>5]) 

# taking out the inside paranthesis piece by piece 
x1>5 
x1[x1>5] 
```

# random example (what is arr.ind argument)
```{r}
# create a matrix
m <- matrix(1:12, 3, 4)
# rownames(m) <- paste("Case", 1:3, sep = "_")
m
which(m %% 5 == 0, arr.ind = TRUE)
which(m %% 5 == 0, arr.ind = FALSE)
```



