---
title: "Plotting with ggplot"
output: html_document
date: "2022-11-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, error = F, message = F, warning = F)
```

# The dataset.

We are using the Palmer Penguins dataset created by Allison Horst. Upload the `palmerpenguins` package and take a look inside the dataset with `glimpse()` function.

```{r}
library(palmerpenguins)
tibble::glimpse(penguins)
```

Before we proceed with plotting, we will remove the `NAs` from the dataset with the `na.omit()` function. We will store the pre-processed data in a variable called `pngs` for the later use. 

```{r}
pngs <- na.omit(penguins)
```

# The default `ggplot`.

Upload the `tidyverse` package to use the functionality of `{ggplot2}`.

```{r}
library(tidyverse)
```

The syntax of `{ggplot}` requires three things to specify:

  * data
  * aesthetics
  * geometry
  
We define the plotting object by calling `ggplot(data=df)` which defines the data we are going to work with. 

We specify the data outside `aes()` and add the variables that `ggplot` maps the aesthetics to inside `aes()`.

We define the positional aesthetics by adding `aes(x=var1, y=var2)` to the `ggplot()` call. 

Next, we define the geometries with, e.g., `geom_point()` to create a scatter plot. There are multiple geometries (starting with `geom_`) one can choose from.


```{r}
(g <- ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm)))
```

In the plot above, the `{ggplot2}` doesn't know yet how we want to plot the data. Let's choose `geom_point()` to create a scatter plot of our data.

```{r}
g + geom_point()
```

# Change the properties of geometries. 

You can manipulate visual aesthetics such as color, shape, and size of points within the `geom_*` command.

You can choose from [preset colors](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf), [hex color codes](https://www.color-hex.com/), both in quotes, or *RGB* colors with `rgb()` function.

```{r}
g + geom_point(color="firebrick", shape="diamond", size=2)
```

# Replace the default `ggplot2` theme.

Let's set a different built-in theme, e.g, `theme_bw()`.

With `theme()` you can modify such elements as texts, rectangles, and lines.

```{r}
theme_set(theme_bw())

g + geom_point(color="firebrick")
```

# Working with axes.

Let's change the axis titles by using `labs()` function.

```{r}
ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm)) + 
  geom_point(color="firebrick") + 
  labs(x="Body Mass", y="Flipper Length (mm)")
```

You can increase space between axis and axis titles with `theme()` function. We can overwrite the default `element_text()` within `theme()` call. `vjust` is the vertical alignment.

```{r}
ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm)) + 
  geom_point(color="firebrick") + 
  labs(x="Body Mass", y="Flipper Length (mm)") + 
  theme(axis.title.x = element_text(vjust = 0, size = 12),
        axis.title.y = element_text(vjust = 2, size = 12))
```

You can also use `margin()` function to adjust the titles. The labels `t` and `r` refer to *top* and *right*.

```{r}
ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm)) + 
  geom_point(color="firebrick") + 
  labs(x="Body Mass", y="Flipper Length (mm)") + 
  theme(axis.title.x = element_text(margin = margin(t = 10), size = 12),
        axis.title.y = element_text(margin = margin(r = 10), size = 12))
```

You can also change the aesthetics of axis titles.

```{r}
ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm)) + 
  geom_point(color="firebrick") + 
  labs(x="Body Mass", y="Flipper Length (mm)") + 
  theme(axis.title = element_text(size = 12, color="firebrick", face="italic"))
```

You can change the aesthetics of the axis text.

```{r}
ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm)) + 
  geom_point(color="firebrick") + 
  labs(x="Body Mass", y="Flipper Length (mm)") + 
  theme(axis.text = element_text(size = 12, color="deepskyblue", face="italic"))
```

You can rotate the axis text by specifying the `angle`. You can adjust the position vertically with `hjust` and horizontally with `vjust`.

```{r}
ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm)) + 
  geom_point(color="firebrick") + 
  labs(x="Body Mass", y="Flipper Length (mm)") + 
  theme(axis.text.x = element_text(angle = 50, vjust = 1, hjust = 1, size = 12))
```

You can limit the axis range, i.e., look closer at a specific data range (without sub-setting the data).

```{r}
ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm)) + 
  geom_point(color="firebrick") + 
  labs(x="Body Mass", y="Flipper Length (mm)") + 
  xlim(c(4000,5000))
```

You can alter your labels.

```{r}
ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm)) + 
  geom_point(color="firebrick") + 
  labs(x="Body Mass", y=NULL) + 
  scale_y_continuous(label=function(x) {return(paste(x, "mm"))})
```

# Working with titles.

We can use `ggtitle()` to add the title.

```{r}
ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm)) + 
  geom_point(color="firebrick") + 
  labs(x="Body Mass", y="Flipper Length (mm)") + 
  ggtitle("Flipper Length in Penguins")
```

```{r}
ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm)) + 
  geom_point(color="firebrick") + 
  labs(x="Body Mass", y="Flipper Length (mm)",
       title="Flipper Length in Penguins",
       subtitle="Flipper length per body mass in penguins",
       caption="Data: Palmer Penguins",
       tag="Fig.1")
```

# Working with legends.

```{r}
ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm, color=species)) + 
  geom_point() + 
  labs(x="Body Mass", y="Flipper Length (mm)")
```

```{r}
ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm, color=species, shape=species)) + 
  geom_point() + 
  labs(x="Body Mass", y="Flipper Length (mm)") +
  guides(color = "none")
```

# Working with facets.

```{r}
g <- ggplot(pngs, aes(x=body_mass_g, y=flipper_length_mm)) + 
  geom_point(color = "chartreuse4", alpha = .3) + 
  labs(x="Body Mass", y="Flipper Length (mm)") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

g + facet_wrap(~species)
```


