---
title: "Tutorial 2 Data Wrangling"
author: "Inga & Lune"
date: "4 11 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# package for convenience functions (e.g. ggplot2, dplyr, etc.)
library("tidyverse")

# remotes::install_github("rstudio/EDAWR")
library(EDAWR)
```

# Motivation

For data scientists, wrangling / manipulating / transforming data is 50 - 80% of time spent -- so we should learn to do this efficiently!


Tidy data can make our lives easier. What are the 3 properties of tidy data? 


### Data wrangling/transformation cheatsheets

https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

https://raw.githubusercontent.com/rstudio/cheatsheets/master/data-transformation.pdf

### Basic R cheatsheet

https://raw.githubusercontent.com/rstudio/cheatsheets/master/base-r.pdf



# Task 1

Experimental Data:

```{r}

data <- 

  tibble(subject_id = rep(c('1', '2', '3'), times = 2),
         Hunger    = rep(c('hungry', 'not_hungry'), each = 3),
         RT = c(100, 300, 120, 110, 90, 100))
data

```

a. Is this a within-subject or between-subject design? 


b. Is this a repeated measure design? 



c. Is this experiment a one-factor or a full factorial design? What is/are the factor(s)? How many levels does each factor have?


d. What is the dependent and what is the independent variable?



# Task 2

Data Wrangling

When looking at this data set, you would like to get the mean RT of each subject. 
You would like to have a tibble with the subject_id and the corresponding mean RT. 


<!-- #region -->

# Now let's look at some datasets which are part of the EDAWR library. 


## First take a look at the storms data
<!-- #endregion -->

```{r}
storms
```

### Split the date column into day, month, and year columns 


### Then, filter out storms which were below 50 in wind speed, and remove the d column. 


## Now let's look at the pollution data

```{r}
pollution
```

### Make new columns using the values from size. It should have city, large, and small columns. Save it as pollution_wider


### Now reverse the operation using pollution_wider. The result should look like the original dataframe.


### Add the city number information to the last column of the pollution table, and sort by amount

Paste the tribble below. Hint: the str_to_title() function might help before joining 


```{r}
city_number  <- tribble(
~city, ~num,
"new york", 1,
"london", 2,
"beijing", 3)

city_number
```



### Group by size and show the average amount for ea. size in a column called mean

## Now let's look at the tuberculosis data from a WHO Global Tuberculosis Report, called tb. 

```{r}
head(tb)
```

### How many countries are there? 



### Now add an extra column called code, which contains the first 3 letters of the country. Move code column next to country column. 



# That's all we have for you today! Any questions? 




