---
title: "Homework Sheet 1 -- R Basics -- Solutions"
date: 'Due: Thursday, November 10 by 23:59 CET'
author: ""
output: 
  html_document:
    toc: true
    toc_depth: 2
    highlight: tango
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, error = F, message = F, warning = F)

```

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}

# package for convenience functions (e.g. ggplot2, dplyr, etc.)
library(tidyverse)

```

# Instructions

* If you need help, take a look at the suggested readings in the lecture, make use of the [cheat sheets](https://rstudio.com/resources/cheatsheets/) and the help possibility in R
* Make sure you have R and RStudio installed. If you are an advanced user and aren't using RStudio, make sure you have everything installed in order to 'knit' the HTML output.
* In the 'author' heading of the Rmd file, write your group members' last names and answer the following questions.

* When all answers are ready, 'Knit' the document to produce a HTML file.

* Create a ZIP archive called "IDA_HW01-Last-names.zip" ('Last-names' are all members in the group as written in the 'author') containing:
   * an R Markdown file "IDA_HW01-Last-names.Rmd"
   * a knitted HTML document "IDA_HW01-Last-names.html"
   * any other files necessary to compile your Rmarkdown to HTML (pictures etc.)
* Upload the ZIP archive on Stud.IP in your group folder before the deadline (see above). You may upload as many times as you like before the deadline, only your final submission will count.

# <span style = "color:firebrick">Exercise 1:</span> Playing around with vectors & matrices


## Ex 1.a [2 points]

Create the vector  [101, 102, 103, 56, 75, 34, 56, 75, 34, 1000, 1100, 1200] using a combination of the c(), seq(), and rep() functions

```{r}
# your code here 
```

**Solution**

```{r}
this_vector <- c(seq(from = 101, to = 103, by = 1), 
                 rep(c(56, 75, 34), 2),
                 seq(from = 1000, to = 1200, by = 100))
this_vector
```

## Ex 1.b [2 points] 

Create the following matrix, which stores the name and suit of every card in a royal flush.

```{r, eval = F}
"ace"   "spades"
"king"  "spades"
"queen" "spades"
"jack"  "spades"
"ten"   "spades"
```


```{r}
# your code here 
```

**Solution**

```{r}
hand1 <- c("ace", "king", "queen", "jack", "ten", "spades", "spades",
      "spades", "spades", "spades")

matrix(hand1, ncol = 2)

```


# <span style = "color:firebrick">Exercise 2:</span> Data frames and tibbles


## Ex 2.a [2 points]

Use the function `tibble` to represent the following information in a tibble:

```{r, eval = F}
title                         director                   year      runtime    berlinale
Metamorphosis of Birds        Catarina Vasconcelos       2020      101        TRUE
Holy Motors                   Leos Carax                 2012      116        FALSE
Kajillionaire                 Miranda July               2020      106        FALSE
Crimes of the future          David Lynch                2022      107        FALSE
Jeanne Dielman                Chantal Akerman            1975      201        FALSE
Italian Studies               Adam Leon                  2021      81         FALSE
```

```{r}
# your code here 
```

**Solution**

```{r}

film_data <- tibble(
    title = c("Metamorphosis of Birds", "Holy Motors", "Kajillionaire","Crimes of the future", "Jeanne Dielman", "Italian Studies"),
    director = c("Catarina Vasconcelos", "Leos Carax", "Miranda July","David Lynch", "Chantal Akerman", "Adam Leon"),
    year = c(2020, 2012, 2020, 2022, 1975, 2021),
    runtime = c(101, 116, 106, 107, 201, 81),
    berlinale  = c(TRUE, FALSE, FALSE, FALSE, FALSE, FALSE)
  )
film_data
```

##  Ex 2.b [2 points]

There is a mistake in the dataset. Rename director of "Crimes of the future" from "David Lynch" to "David Cronenberg" 

```{r}
# your code here 
```

**Solution**

```{r}
#film_data[4,2] <- "David Cronenberg"
film_data$director[4] <- "David Cronenberg"
film_data
```

## Ex 2.c [2 points]

show the unique values of the column 'year'. 

```{r}
# your code here
```

**Solution**

```{r}
film_data$year |> 
  unique()
```


# <span style = "color:firebrick">Exercise 3:</span> Functions and piping


## Ex 3.a [3 points] 

Fill in the gaps to define a function that simulates rolling a pair of weighted dice:

You will need to add a `prob` argument to the sample function inside of the `roll` function. This argument should be a vector of weighted probabilities so that the number 6 comes up with probability 3/8, and every other number with probability 1/8.

```{r}
roll <- function() {
  die <- ...
  dice <- sample(die, size = ..., replace = ..., ...)
  sum(dice)
}
```

**Hint:** Check the `sample` function. 

**Solution**

```{r}
roll <- function() {
  die <- 1:6
  dice <- sample(die, size = 2, replace = TRUE,
  prob = c(1/8, 1/8, 1/8, 1/8, 1/8, 3/8))
  sum(dice)
}

```

```{r}
# To visualize the long-term behavior of your dice
# You just need to run the code chunk, nothing to add :) 
rolls <- replicate(10000, roll())
qplot(rolls, 
      binwidth = 1,
      xlab = "Sum of the two dice throws", 
      ylab = "Frequencies", 
      fill = I("tomato4"),
      col = I("black")) +
  scale_x_continuous(breaks = 2:12)

```

## Ex 3.b [3 points]

Consider this vector of weirdly specified reaction times (similar to Exercise 2.14 from the web-book). 

```{r}
weird_RTs <- c("RT = 323", "ReactTime = 345", "howfast? -> 23 (wow!)", "speed = 421", "RT:50")
```

Starting with that vector use a chain of pipes to: 

- extract the numeric information from the strings (using stringr functions), 
- cast the information into a vector of type `numeric`, 
- remove all RT entries lower than 100
- take the mean. 

**Hint:** Check the use of regular expressions in the [cheat sheet of the `stringr` package](https://raw.githubusercontent.com/rstudio/cheatsheets/main/regex.pdf). 

```{r}
# your code here
```

**Solution:**

```{r}
weird_RTs |> 
  str_extract("(\\d)+")  |>  
  as.numeric() |> 
  (function(x) {x[x>100]})() |> 
  mean()

# Another way to only extract numbers from string vector (not part of solution)
gsub("[^0-9]", "", weird_RTs) |> 
  as.numeric()

```

## Ex 3.c [2 points] 

A friend gives you some useful information in a useless format (a standard problem of data analysis):

```{r}
norse_myth <- "Baldr %% Odin&Frigg->Loki _io of Hel* Fenrir* and Jörmungandr."

```

Use piping and some magic from the `stingr` package to produce output  like this:

```{r, eval = F}
baldr is the child of odin&frigg.
loki is the father of hel, fenrir, and jörmungandr.

```

[**Hints:** 1.Use of `"\n"` inserts a line break, but it shows only when using `cat` to print the output. 2. pay attention to capital / lower letters]

```{r}
# your code here
```

**Solution:**

```{r}
norse_myth  |> 
  str_replace_all(pattern = "->", replacement = ".\n") |> 
  str_replace(pattern = "%%", replacement = "is the child of") |> 
  str_replace_all(pattern = "\\*", replacement = ",")  |> 
  str_replace(pattern = "_io", replacement = "is the father")  |> 
  tolower() |> 
  cat()
```
