---
title: "Homework Sheet 1 -- R Basics -- Solutions"
date: 'Due: Thursday, November 4 by 11:59 CET'
author: ""
output: 
  html_document:
    toc: true
    toc_depth: 2
    highlight: tango
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, error = F, message = F, warning = F)

```

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}

# package for convenience functions (e.g. ggplot2, dplyr, etc.)
library(tidyverse)

```

# Instructions

* If you need help, take a look at the suggested readings in the lecture, make use of the [cheat sheets](https://rstudio.com/resources/cheatsheets/) and the help possibility in R
* Make sure you have R and RStudio installed. If you are an advanced user and aren't using RStudio, make sure you have everything installed in order to 'knit' the HTML output.
* Create an Rmd-file with your group number (equivalent to StudIP group) in the 'author' heading and answer the following questions.
* When all answers are ready, 'Knit' the document to produce a HTML file.
* Create a ZIP archive called "IDA_HW01-Group-XYZ.zip" (where 'XYZ' is *your* group number) containing:
   * an R Markdown file "IDA_HW1-Group-XYZ.Rmd"
   * a knitted HTML document "IDA_HW01-Group-XYZ.html"
   * any other files necessary to compile your Rmarkdown to HTML (pictures etc.)
* Upload the ZIP archive on Stud.IP in your group folder before the deadline (see above). You may upload as many times as you like before the deadline, only your final submission will count.

# <span style = "color:firebrick">Exercise 1:</span> Playing around with vectors & matrices

## Ex 1.a [2 points]

Create a named Boolean vector and assign it to variable `passed`. This vector should contain the following information about who passed and didn't pass the final exam: Jax passed, Jamie passed, Jason didn't, Geralt did. 

```{r}
# your code here 
```

**Solution**

```{r}
passed <- c(Jax = T, Jamie = T, Jason = F, Geralt = T)
```

## Ex 1.b [2 points]

Compute the proportion of participants who passed by applying a single function `XXX` to the vector `passed`, like `XXX(passed)`:

```{r}
# your code here 
```

**Solution**

```{r}
mean(passed)
```

## Ex 1.c [2 points]

Use an appropriate function from the `stringr` package to obtain a Boolean vector indicating whether a name of the entries in `passed` (i.e., a name of the participants in our class) starts with the letter "J". Use this Boolean vector to obtain the subset of entries from `passed` for all participants whose name starts with the letter "J".

```{r}
# your code here 
```

**Solution**

```{r}
passed[str_detect(names(passed), "^J")]
```

## Ex 1.d [4 points]

Use a single line of R code to return a matrix with the following values by using `seq` and piping `%>%`.

```{r, eval = F}
 3   9    15 
 5   11   17
 7   13   19
```

```{r}
# your code here
```

**Solution:**

```{r}
seq(from = 3, to = 19, by = 2) %>% matrix(nrow = 3)
```


# <span style = "color:firebrick">Exercise 2:</span> Data frames and tibbles

## Ex 2.a [2 points]

Use the function `tribble` to represent the following information in a tibble:

```{r, eval = F}
participant   condition     response
S01           A             32
S01           B             54
S02           A             23
S02           B             03
```

```{r}
# your code goes here
```

**Solution:**

```{r}
tribble(
~participant,   ~condition,     ~response,
"S01",           "A",             32,
"S01",           "B",             54,
"S02",           "A",             23,
"S02",           "B",             03
)
```


## Ex 2.b [2 points]

Create the exact same tibble, but now using the command `tibble`

```{r}
# your code goes here
```

**Solution:**

```{r}
tibble(
  participant = c("S01", "S01", "S02", "S02"),
  condition   = c("A", "B", "A", "B"),
  response    = c(32, 54, 23, 3)
)
```

## Ex 2.c [2 points]

Which (if any) of these columns could reasonably be represented as a factor?

... your answer here ... 

**Solution:**

Whatever the "response" is, it is likely a numerical measure, so not something that falls in a small set of relevant categories. But there are finitely many participants and conditions, so it might make sense to treat these columns as factors.

## Ex 2.d [4 points]

Create a tibble with data from a larger imaginary experiment. Concretely, your tibble should have 100 rows, and four columns called `participant`, `trial`, `condition` and `response`. There are five conditions, labelled A, B, C, D and E. There are ten participants, each of which sees all five conditions exactly twice. Participants are represented with strings like "S1", "S2" and so on. You can make up random real numbers sampled uniformly at random from 0 to 100 (rounded to two digits after the comma) to fill the `response` column. Below, we've already started on the code, but some of our code is wrong, some parts might give good results but are really clumsy and inelegant. Finish the job and improve where you can! (Check out help pages for all commands you do not know.)

```{r, eval = F}
exp_data <- tibble(
  participant = str_c("", seq(from = 1, to = 100, by = 1), sep = "S"),
  condition   = rep(1:10, each = 10), 
  trial       = c("A", "B", "C", "D", "E"),
  response    = (runif(100) - 5) %>% ceiling()
)
```

**Solution:**

```{r, eval = F}
exp_data <- tibble(
  participant = str_c("S", 1:10) %>% rep(each = 10),
  condition   = c("A", "B", "C", "D", "E") %>%  rep(each = 2) %>% rep(times = 10),
  trial       = rep(1:10, times = 10), 
  response    = (runif(100) * 100) %>% round(2)
)
exp_data
```

# <span style = "color:firebrick">Exercise 3:</span> Functions and piping

## Ex 3.a [4 points]

A friend gives you some useful information in a useless format (a standard problem of data analysis):

```{r}
great_info_in_crap_format <- "Johnny_Rotten->Sex_Pistols*Johnny_Ramone->Ramones*Johnny_Cash->The_Tennessee_Three"
```

Use piping and some magic from the `stingr` package to produce output (as close as possible to something) like this:

```{r, eval = F}
Johnny Rotten was part of the Sex Pistols.
Johnny Ramone was part of the Ramones.
Johnny Cash played with The Tennesse Three.
```

[**Hints:** Use of `"\n"` inserts a linebrake, but it shows only when using `cat` to print the output.]

```{r}
# your code here
```

**Solution:**

```{r}
great_info_in_crap_format %>%
  str_c("*") %>% 
  str_replace_all(pattern = "_", replacement = " ") %>% 
  str_replace_all(pattern = "\\*", replacement = ".\n") %>% 
  str_replace(pattern = '->The Tennessee Three', replacement = ' played with The Tennessee Three') %>% 
  str_replace_all(pattern = '->', replacement = ' was part of the ') %>% 
  cat()
```

## Ex 3.b [4 points]

Write a named function called `even_cumulative_mean` that takes a numeric vector `input` as argument. The function should first check whether the input is numeric (using `is.numeric`) and whether it has more than one element. If these conditions are not fulfilled, informative error messages should be given (with the `stop` command). If the input check succeeds, it should return a tibble with three columns (choose appropriate names yourself):

1. the list of all even indices for which entries exist in `input`,
2. all of the entries in `input` with an even index
3. a vector whose $i$th entry is the mean over all entries in the second column up to index $i$

Apply your function to the following vectors:

```{r}
input_1 <- c(12, 43, 56, 87, 98)
input_2 <- c(12, 43, 56, 87, 98, 101)
```

```{r}
# your code here
```

**Solution:**

```{r}
even_cumulative_mean <- function(input) {
  if (! is.numeric(input)) {
    stop("Input to 'odd_cumulation' should be numeric.")
  }
  if (length(input) <= 1) {
    stop("Input to 'odd_cumulation' must have at least two elements")
  }
  last_even_index <- ifelse(length(input) %% 2 == 0, length(input), length(input)-1)
  even_numbers <- seq(from = 2, to = last_even_index, by = 2)
  tibble(
    even_numbers = even_numbers,
    input_at_even = input[even_numbers],
    cumulative_mean = (map_dbl(1:length(input_at_even), ~ input_at_even[1:.x] %>% mean))
  ) 
}

even_cumulative_mean(input_1)
even_cumulative_mean(input_2)
```

## Ex 3.c [4 points]

Consider this vector of weirdly specified reaction times (similar to Exercise 2.14 from the web-book). 

```{r}
weird_RTs <- c("RT = 323", "ReactTime = 345", "howfast? -> 23 (wow!)", "speed = 421", "RT:50")
```

Starting with that vector use a chain of pipes to: 

- extract the numeric information from the strings, 
- cast the information into a vector of type `numeric`, 
- remove all RT entries lower than 100
  - if you can, use an anonymous function defined in situ; otherwise define a named function; 
  - if you can, use Booleans and indexing, not some other other ready-made function, 
- take the log, take the mean, 
- round to 3 significant digits

**Hint:** Check the use of regular expressions on the [cheat sheet of the `stringr` package](https://github.com/rstudio/cheatsheets/raw/master/strings.pdf). 

```{r}
# your code here
```

**Solution:**

```{r}
weird_RTs %>% 
  stringr::str_extract("[:digit:]+") %>% 
  as.numeric() %>% 
  (function(x) {x[x>100]}) %>% 
  log %>% 
  mean %>% 
  signif(3)
```




